---
title: "graduation.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8252.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in [graduation.csv](https://github.com/zief0002/epsy-8252/raw/master/data/graduation.csv) include student-level attributes for $n=2344$ randomly sampled students who were first-year, full-time students from the 2002 cohort at a large, midwestern research university. Any students who transferred to another institution were removed from the data. The source of these data is: @Jones-White:2014. The attributes, collected for these students are:

- `degree`: Dummy-coded variable to indicate whether the student graduateg from the institution (0 = No; 1 = Yes)
- `act`: Student's ACT score (If the student reported a SAT score, a concordance table was used to transform the score to a comparable ACT score.)
- `scholarship`: Amount of scholarship offered to student (in thousands of dollars)
- `ap`: Number of Advanced Placement credits at time of enrollment
- `firstgen`: Dummy-coded variable to indicate whether the student was a first generation college student? (0 = No; 1 = Yes)
- `nontrad`: Dummy-coded variable to indicate whether the student was a non-traditional student (older than 19 years old at the time of enrollment)? (0 = No; 1 = Yes)



### Preview

```{r}
options(dplyr.width = Inf, width = 85)
grad = readr::read_csv(file = "~/Documents/github/epsy-8252/data/graduation.csv")
head(grad)
```


### References




