---
title: "movies.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in [movies.csv](https://github.com/zief0002/epsy-8251/raw/master/data/movies.csv) includes attributes for $n=292$ movies. These data are a subset of data from the `movies` data object included in the **ggplot2movies** package. The original data contains information on 24 variables collected from 28,819 movies. The attributes in this sample include::

- `title`: Movie's title
- `budget`: Movie's budget (in millions of U.S. dollars); adjusted to 2018 dollars
- `year`: Release date of the movie
- `age`: Age of the movie; Computed by subtracting the movie's release date from 2019
- `genre`: Movie genre (Action, Comedy, Drama, Other)
- `length`: Running time of the movie (in minutes)


### Preview

```{r}
movies = readr::read_csv(file = "~/Documents/github/epsy-8252/data/movies.csv")
head(movies)
```


### References

<p style="padding-left: 30px; text-indent: -30px;">Wickham, H. (2015). *ggplot2movies: Movies data.* R package version 0.0.1.
  https://CRAN.R-project.org/package=ggplot2movies</p>
