---
title: "nba-team-data.csv and nba-player-data.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8252.bib'
csl: '../../notes/style/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in *nba-team-data.csv* and *nba-player-data.csv*, inspired by @Woltman:2012, include player-level attributes for 300 NBA players, and team-level attributes for 30 different teams, respectively. 

The player-level attributes in *nba-player-data.csv* include:

- `player:` Name of the NBA player
- `team:` Name of the NBA team for each player
- `success:` A proxy for player quality/success. This is the quantile for the player based on the player's free-throw percentage - relative to the other players in the league. Higher values indicate a more successful player (e.g., 0 = lowest 20%; 4 = highest - 20%).
- `life_satisfaction:` Score on a survey of life satisfaction. Scores range from 5 to 25, with higher scores indicating more life satisfaction.

The team-level attributes in nba-team-data.csv include:

- `team:` Name of the NBA team
- `coach:` Name of the team's current coach
- `coach_experience:` This is the tercile for the coach based on years of coaching experience in the NBA (e.g., 0 = lowest third; 2 = highest third).


### Preview



**NBA team data**

```{r}
# options(dplyr.width = Inf, width = 85)
nba_team = readr::read_csv(file = "~/Documents/github/epsy-8252/data/nba-team-data.csv")
head(nba_team)
```

*NBA players data**

```{r}
# options(dplyr.width = Inf, width = 85)
nba_player = readr::read_csv(file = "~/Documents/github/epsy-8252/data/nba-player-data.csv")
head(nba_player)
```


### References

<!-- <p style="padding-left: 50px; text-indent: -50px;"> -->

