---
title: "nels.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This data, stored in [nels.csv](https://github.com/zief0002/epsy-8252/raw/master/data/nels.csv) includes three attributes on $n = 744$ 10th-grade students. These data come from @Keith:2015. The attributes are:

- `social10`: 10th-grade standardized achievement score on History/Civics/Georgraphy (Social Studies) test
- `self_esteem`: Standardized measure of self-esteem; administered in 10th-grade
- `locus`: Standardized measure of locus-of-control; more positive values indicate higher internal locus-of-control
- `ses`: Standardized measure of socio-economic status, including parent education level
- `gpa`: Grade-point average (GPA) from grades6&ndash;8 in all subjects (on a standard 4-point scale)
- `parent_ed`: Education-level (in years of schooling) for the parent with the highest level of education


### Preview

```{r}
options(dplyr.width = Inf, width = 85)
nels = readr::read_csv(file = "~/Documents/github/epsy-8252/data/nels.csv")
head(nels)
```


### References

<!-- <p style="padding-left: 50px; text-indent: -50px;"> -->

