---
title: "netherlands-schools.csv and netherlands-students.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8252.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in *netherlands-students.csv* and *netherlands-schools.csv* include student- and school-level attributes, respectively, for $n_i=2287$ 8th-grade students from $n_j = 131$ schools in the Netherlands provided by @Snijders:2012.

The student-level attributes in *netherlands-students.csv* include:

- `school_id`: The school ID number for each student
- `language_pre`: Language pre-test score
- `language_post`: Language post-test score
- `ses`: Measure of the socio-economic status
- `verbal_iq`: Student's score on a verbal IQ test. The variable is centered to have a mean of 0. 
- `female`: Student's sex (0 = male; 1 = female)
- `minority`: Student's minority status (0 = white; 1 = minority)

The school-level attributes in *netherlands-schools.csv* include:

- `school_id`: The school ID number
- `school_type`: Indicates whether the school is a public school, a Protestant private school, a Catholic private school, or a non-denominational private school
- `public`: Indicates whether the school is a public school (1) or a private school (0)
- `school_ses`: School's average socio-economic status
- `school_verbal_iq`: School's average verbal IQ score
- `school_minority`: Percentage of students at the school who are minority students






### Preview

**Student data**

```{r}
# Read in student-level data
student_data = readr::read_csv(file = "~/Documents/github/epsy-8252/data/netherlands-students.csv")
head(student_data)
```

**School data**

```{r}
# Read in school-level data
school_data = readr::read_csv(file = "~/Documents/github/epsy-8252/data/netherlands-schools.csv")
head(school_data)
```



### References



