---
title: "popular-classroom.csv and popular-student.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8252.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in *popular-classroom.csv* and *popular-student.csv* includes data on 2000 different students from 100 different classrooms. The data, provided by @Hox:2002, were simulated from data collected as part of a sociological study of student popularity. Student popularity, a rating on a scale of 1&ndash;10, was derived by a sociometric procedure in which all students in a class rate all the other students. Each students' popularity is the average received popularity rating.  

The attributes in *popular-classroom.csv* include:

- `class`: Classroom ID
- `teacherExp`: Experience level of teacher, in years

The attributes in *popular-student.csv* include:

- `student`: Student ID (within a school)
- `class`: Classroom ID 
- `popularity`: Popularity rating of the student (0-10 scale; where 0 = very unpopular and 10 = very popular) based on having all other students in the student's class rate him/her and computing the average of those ratings.
- `teachPop`: Student's popularity as rated by the teacher. Higher values are indicative of higher popularity
- `extra`: Extraversion of the student (measured on a 10-point scale)
- `female`: Sex of the student (Male = 0; Female = 1)


### Preview



**Classroom data**

```{r}
# options(dplyr.width = Inf, width = 85)
classroom = readr::read_csv(file = "~/Documents/github/epsy-8252/data/popular-classroom.csv")
head(classroom)
```

**Student data**

```{r}
# options(dplyr.width = Inf, width = 85)
student = readr::read_csv(file = "~/Documents/github/epsy-8252/data/popular-student.csv")
head(student)
```


### References

<!-- <p style="padding-left: 50px; text-indent: -50px;"> -->

