---
title: "vocabulary.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8252.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data, adapted from data provided by @Bock:1975, come from the Laboratory School of the University of Chicago and include scaled test scores across four grades from the vocabulary section of the Cooperative Reading Test for $n=64$ students. The attributes in the dataset include:

- `id`: The student ID number
- `vocab_08`: The scaled vocabulary test score in 8th grade
- `vocab_09`: The scaled vocabulary test score in 9th grade
- `vocab_10`: The scaled vocabulary test score in 10th grade
- `vocab_11`: The scaled vocabulary test score in 11th grade
- `female`: Dummy coded sex variable (0 = Not female, 1 = Female)


### Preview

```{r}
options(dplyr.width = Inf, width = 85)
vocab = readr::read_csv(file = "~/Documents/github/epsy-8252/data/vocabulary.csv")
head(vocab)
```


### References


