---
title: "Untitled"
author: "Andrew Zieffler"
date: "1/7/2021"
output: 
  html_document:
    highlight: zenburn
bibliography: "myBibliography.bib"
csl: "apa-single-spaced.csl"
---

The hat notation (e.g., $\hat{Y_{ij}}$) is used to indicate an estimate.



# Adding Citations

Here is some text and a citation [@carmichael_laziness_1954].

In this example, we have prefixed the citation with the word "see" and added a page number after the citation by including "p. 208". The identifier and the text following the identifier are separated by a comma. [see @carmichael_laziness_1954, p. 208].

@carmichael_laziness_1954 suggest that something is true. 

Add multiple references by including multiple citation identifiers separated by a semicolon [e.g., @ross_racial_2020; @carmichael_laziness_1954]


# Syntax

```{r}

```


# Table Creation

```{r table_01}
library(knitr)
library(kableExtra)
library(tidyverse)

# Input cell information
tab_01 = data.frame(
  scale = c("BAS-T", "SR", "BDI", "ASRM", "M-SRM"),
  high = c("46.17 (2.87)", "17.94 (1.88)", 
           "7.11 (6.50)", "6.46 (4.01)", "11.05 (3.36)"),
  mod = c("37.99 (1.32)", "11.52 (1.84)", 
          "6.18 (6.09)", "5.63 (3.69)", "11.76 (2.75)"),
  p = c("<.001", "<.001", ".254", ".109", ".078")

)

# Show output
tab_01 %>%
  kable(
    format = "html",
    col.names = c("Scale", "High BAS group", "Moderate BAS group", "*p*"),
    caption = "**Table 1** <br /> *Means and Standard Deviations of Scores on Baseline Measures*",
    align = c("l", "c", "c", "c"),
    table.attr = "style='width:40%;'"
  ) %>%
  kable_classic(html_font = "\"Helvetica Neue\"") %>%
  row_spec(row = 0, align = "c") %>%
  footnote(
    general = "Standard deviations are presented in parentheses. BAS = Behavioral Activation System; BAS-T = Behavioral Activation System-Total scores from the Behavioral Inhibition System/Behavioral Activation System Scales; SR = Sensitivity to Reward scores from the Sensitivity to Punishment and Sensitivity to Reward Questionnaire; BDI = Beck Depression Inventory scores; ASRM = Altman Self-Rating Mania Scale scores; M-SRM = Modified Social Rhythm Metric Regularity scores.",
    general_title = "Note.",
    footnote_as_chunk = TRUE
    )
```

# References